---
title: "Liuda R tutorial"
author: "O.M."
date: "2024-11-23"
output: html_document
---


Lets begin from the beggining
```{r}
print("Hello World")
```
Here Print your command
```{r}

```


##Objects##
Then lets go to a type of objects
First one is numerical - numbers
```{r}
1
typeof(1)
```

```{r}
1/2
1.5
typeof(1/2)
typeof(1.5)
```
Integers are doubles but without decimal
```{r}
typeof(3.14)
as.integer(3.14) 
```


Now to strings - simplest string is a character - or a bunch of them
String should be separated by either "" or '', and represent line of text whihc shouldnt be treates as number
```{r}
"Hello"
typeof("Hello")
```
Numeric in "" brackets would also be treated as character
```{r}
"2"
typeof("2")
```

Last type is a factor
Factor is type of character whihc can have only some specuific values 
c() fucntion is used to concatenate any objects in one string
```{r}
1
factor(c(1, 2))
```

```{r}
c(1, 4, 7, 1)
```

```{r}
c("f", "u", "c", "k")
```
You can concatenate any type of data
```{r}
c(1, 2, "g")
```
But it would be treated as a character if at least one character is there
```{r}
typeof(c(1, 2, "g"))
typeof(c(1, 2))
```


What would be a type of  next expressions:
one
_
i4u
try them here
```{r}

```


#Operators and indexes#
Operators in R are mostly the same as in other languages 
```{r}
1 + 1

```

```{r}
1 - 1
```

```{r}
2/4
```

```{r}
2*4
```
square root
```{r}
sqrt(4)
```

```{r}
2^2
```
logarithm on base 2
```{r}
log(2)
```
logarithm on base 10
```{r}
log10(2)
```
trigonometric
```{r}
cos(2)
sin(2)
tan(2)
```
For comparison you can use classic >, < =>, <=, and == for checking if two values are equal
```{r}
1>2
2>=2
"two"=="two"
```




To assign value to a varible you can use either = or <- / ->
But usage of = for assigemnt is disregarded
```{r}
a<-1
a
```


```{r}
2->b
b
```


```{r}
c=3
c
```
Please write a compariosn where you took a sum of logartyhm value of 5 and 4 in 5 degree and check if it is more than square root of 10900
Then assign result of this test to a variable and check its class
```{r}

```




More complex objects
To get acces to a certain element of a list you need to use [] brackets
Important - in R we are starting to count from 1 not from 0
```{r}
our_list<-c(10:20)#using : you will get sequantialy all numbers between first and second
our_list[2]
```

