---
title: "Liuda R tutorial"
author: "O.M."
date: "2024-11-23"
output: html_document
---


```{r}
print("Hello World")
```

Exercise:
Replace "Hello World" with your name or a short greeting of your choice.
```{r}

```

Objects in R
Numerical Data
Numerical values are the most basic data type in R.

```{r}
1
typeof(1)
```

Perform arithmetic operations:
```{r}
1 / 2
1.5
typeof(1/2)
typeof(1.5)
```
Note: Integers are a subset of doubles, but without decimals. Convert a double to an integer:

```{r}
as.integer(3.14)
```


Exercise:
Create and manipulate numerical data, then check its type.
```{r}

```

Character Data (Strings)
Strings represent text, enclosed in double " or single ' quotes.


```{r}
"Hello"
typeof("Hello")
```

Even numbers in quotes are treated as characters:
```{r}
"2"
typeof("2")
```

Exercise:
Write your favorite quote as a string and check its type.
```{r}

```

Factors
Factors are categorical variables with predefined levels.

```{r}
factor(c("red", "blue", "green"))
```

Concatenate using c():
```{r}
c(1, 4, 7)
c("data", "science", "R")
```

```{r}
c(1, 2, "analysis")
typeof(c(1, 2, "analysis"))
```


Create a factor with days of the week.
Mix numbers and characters in a concatenation.
```{r}

```



Variables in R
Variables store data and can be assigned using <-, ->, or =.
```{r}
x <- 5
x

5 -> y
y

b = "Hello"
b
```

Rules for Variable Names
Must start with a letter.
Can contain letters, numbers, dots, or underscores.
Cannot contain spaces or start with a number.

```{r}
valid_variable <- 10
anotherValid.variable <- 20
```

Exercise:
Create variables with valid names storing your age, favorite number, or text.
Try creating a variable with an invalid name and observe the error.
```{r}

```


Structured and Unstructured Data
Structured Data
Structured data is organized into rows and columns, like in a table or dataframe.

```{r}
df <- data.frame(
  Name = c("Alice", "Bob", "Charlie"),
  Age = c(25, 30, 35),
  Score = c(90, 85, 88)
)
df
```
Access elements:
```{r}
df$Name      # Access the 'Name' column
df[1, ]      # Access the first row
df[1, "Age"] # Access the 'Age' value of the first row
```

Exercise:
Create a dataframe with your own data (e.g., Name, Hobby, Favorite Number).
Access specific rows or columns.
```{r}

```

Unstructured Data
Unstructured data lacks a predefined format, like text, images, or raw sensor data.
```{r}
text <- "R is great for data analysis and visualization!"
nchar(text)  # Count characters
strsplit(text, " ")  # Split into words
```

Exercise:
Write a sentence as unstructured data. Count the words in it.
Split the text into individual words and analyze their length with len() function.
```{r}

```

Comparing Structured and Unstructured Data
```{r}
structured <- data.frame(
  Category = c("A", "B", "C"),
  Value = c(10, 20, 30)
)
structured

unstructured <- "Category A has 10 units, B has 20, and C has 30."
unstructured
```



Data frames
Working with Different Types of Variables in a Dataframe
A dataframe in R is a structured collection of data where each column can store a different type of variable (numerical, character, factor, etc.).

Creating a Dataframe with Multiple Variable Types
You can combine variables of different types into a dataframe:
```{r}
df <- data.frame(
  Name = c("Alice", "Bob", "Charlie"),  # Character (categorical)
  Age = c(25, 30, 35),                 # Numeric
  Score = c(90.5, 85.3, 88.9),         # Double
  Passed = c(TRUE, TRUE, FALSE)        # Logical
)
df
```


Checking Variable Types
To check the types of variables in a dataframe, use the str() or sapply() functions:
```{r}
str(df)          # Overview of dataframe structure
sapply(df, typeof) # Check each column's type
```



Accessing Columns
You can access specific columns by their names or indices:
```{r}
df$Name     # Access 'Name' column
df[["Age"]] # Another way to access 'Age' column
df[, 3]     # Access the 3rd column
```



Exercise:
Access the Score column and calculate the average.
Access the Passed column and count how many TRUE values are present.
```{r}

```



Mixing Types of Variables
Numerical Variables
Columns with numbers can store integer or double values. Operations can be applied directly:

```{r}
mean(df$Age)      # Calculate the mean of 'Age'
sum(df$Score)     # Sum of 'Score'
df$Age + 5        # Add 5 to all ages


df
```


Character Variables
Character columns store text and are treated as categorical by default in R dataframes:
```{r}
toupper(df$Name)  # Convert names to uppercase
nchar(df$Name)    # Count the number of characters in each name
```


Factor Variables
Convert character variables to factors to work with specific categories:
```{r}
df$Name <- as.factor(df$Name)
levels(df$Name)  # Check levels of the factor
```

Logical Variables
Logical columns store TRUE/FALSE values, useful for filtering:
```{r}
df$Passed[df$Passed == TRUE]   # Filter rows where 'Passed' is TRUE
```


Adding and Modifying Variables in a Dataframe
Adding a New Column
You can add new variables to the dataframe, derived from existing ones:
```{r}

df$Grade <- ifelse(df$Score > 88, "A", "B")  # Add a grade column
df
```

Changing Variable Types
If needed, change the type of a column:

```{r}
df$Age <- as.character(df$Age)  # Convert 'Age' to character
typeof(df$Age)
```


Dealing with Mixed Variable Types
Converting Columns
R automatically tries to unify types in mixed columns, often converting numbers to characters:

```{r}
mixed <- c(1, "two", 3)
typeof(mixed)  # Will be character
```


Handling NA (Missing Values)
Missing values (NA) can appear in any variable type. Handle them using functions like is.na():
```{r}
df$Age[2] <- NA     # Introduce an NA value
mean(df$Age, na.rm = TRUE)  # Calculate mean ignoring NA
```


Exercises
Create a Dataframe:
```{r}

```

Create a dataframe with at least three columns:
One numeric (e.g., age).
One character (e.g., name).
One logical (e.g., likes ice cream).
```{r}

```



Modify the Dataframe:
Add a new column that calculates the age in months.
Convert one column to a different type (e.g., Name to a factor).
```{r}

```


Filter Rows:
Filter the dataframe to show only rows where LikesIceCream is TRUE.
Summarize Variables:
```{r}

```


DPplyr/tidyverse

The dplyr package from the tidyverse ecosystem is a powerful tool for working with dataframes in R. Let's learn some basic operations.

Loading the dplyr Package
Before you begin, make sure to load the package
If you don’t have it installed, you can install it first
```{r}
library(dplyr)
#install.packages("dplyr")
```

Basic Operations
1. Selecting Columns
Use the select() function to choose specific columns from a dataframe.

```{r}
df <- data.frame(
  Name = c("Alice", "Bob", "Charlie"),
  Age = c(25, 30, 35),
  Score = c(90.5, 85.3, 88.9),
  Passed = c(TRUE, TRUE, FALSE)
)

# Select specific columns
select(df, Name, Age)
# Exclude specific columns
select(df, -Passed)
```


Exercise:
Select only the Score and Passed columns from the dataframe.
```{r}

```

2. Filtering Rows
Use filter() to select rows based on conditions.

```{r}
filter(df, Age > 30)
filter(df, Passed == TRUE)
```

Exercise:
Filter rows where Score is greater than 88.
```{r}

```


3. Adding New Columns
Use mutate() to add or modify columns.
```{r}
mutate(df, Grade = ifelse(Score > 88, "A", "B"))

mutate(df, AgeMonths = Age * 12)
```


Create a new column that calculates the square root of the Score.
```{r}

```

4. Summarizing Data
Use summarize() to calculate summary statistics.

```{r}
summarize(df, AverageScore = mean(Score))
summarize(df, AverageScore = mean(Score), TotalPassed = sum(Passed))
```

To summarize groups of data, combine summarize() with group_by():
```{r}

df %>%
  group_by(Passed) %>%
  summarize(
    AverageScore = mean(Score),
    Count = n()
  )
```


Exercise:
Group by Grade and calculate the total number of people in each grade category.
```{r}

```

5. Rearranging Rows
Use arrange() to sort rows.
```{r}
arrange(df, Score)
```

# Sort rows by Score in descending order
```{r}
arrange(df, desc(Score))
```


Exercise:
Sort the dataframe by Age in ascending order.
```{r}

```

Combining Operations with the Pipe (%>%)
The pipe operator %>% makes your code more readable by chaining operations together.
```{r}
df %>%
  select(Name, Score) %>%
  filter(Score > 88) %>%
  arrange(desc(Score))

```
Exercise:
Select the Name and Age columns.
Filter rows where Age is greater than 28.
Add a column for age in months.
Arrange the result by Age in descending order.
```{r}

```



**select()**	Choose specific columns.
**filter()**	Filter rows based on conditions.
**mutate()**	Add or modify columns.
**summarize()**	Summarize data (e.g., mean, sum).
**group_by()**	Group data for grouped summaries.
**arrange()**	Sort rows in ascending/descending order.

If and loops
Conditional Statements: if and ifelse
1. if Statement
The if statement is used to execute code only if a condition is TRUE.
```{r}
x <- 5
# Check if x is greater than 3
if (x > 3) {
  print("x is greater than 3")
}
```

You can include an else block to handle the case when the condition is FALSE:
```{r}
x <- 2

if (x > 3) {
  print("x is greater than 3")
} else {
  print("x is less than or equal to 3")
}
```

Exercise:
Write an if statement to check if a number y is positive, negative, or zero, and print an appropriate message.
```{r}

```

2. ifelse() Function
The ifelse() function is a vectorized version of if and is great for conditional logic in vectors.

```{r}
x <- c(1, -2, 3, -4, 5)
# Check if elements are positive or negative
ifelse(x > 0, "Positive", "Negative")
```


You can also assign the result to a variable:
```{r}
status <- ifelse(x > 0, "Positive", "Negative")
status
```

Exercise:
Create a vector z <- c(10, 0, -10, 5, -5) and use ifelse() to label each number as "Positive", "Zero", or "Negative".
```{r}

```

Loops: for and while
3. for Loops
A for loop iterates over a sequence.
```{r}
# Print numbers from 1 to 5
for (i in 1:5) {
  print(i)
}

```
# Loop through a vector
```{r}
fruits <- c("Apple", "Banana", "Cherry")
for (fruit in fruits) {
  print(fruit)
}
```

Exercise:
Write a for loop to print the square of each number in the vector c(1, 2, 3, 4, 5).
```{r}

```

4. while Loops
A while loop continues as long as a condition is TRUE.
```{r}
# Print numbers from 1 to 5
i <- 1
while (i <= 5) {
  print(i)
  i <- i + 1
}
```

Exercise:
Write a while loop to print all even numbers from 10 down to 2.
```{r}

```


